Efficient Thresholded Correlation using Truncated Singular Value Decomposition

### 2015-07-28 threshold/draft.Rnw
Typos; some more detail about other methods

The number of correlation coefficients grows quadratically with the number of
vectors. Simply computing all pairs of correlation coefficients and then
filtering out coefficients below a given threshold may not be computationally
feasable for high-dimensional data arising in modern genomics and other
applications. Considerable attention has been devoted to methods called
\emph{pruning methods} that cheaply prune away all but the most
highly-correlated pairs of vectors.  A variety of pruning methods have been
described for the thresholded correlation problem; see, for
example\cite{prune1, prune2} and the references therein. Related methods
approximate a set containing the most highly-correlated pairs\cite{prune3,
timeseries}. A method of Wu, et al.\cite{svd-similarity}, conceptually very
similar to our approach, uses the singular value decomposition (SVD) to find
pairs of close vectors with respect to a distance metric (instead of
correlation).  Our method described below prunes along an intuitive path of
decreasing variance defined by the data and computed by the SVD.

Consider a real-valued data matrix $A$ consisting of $m$ observations of $n$
column vectors.  Denote the columns of $A$ as $a_j\in\mathcal{R}^m$ for
$j=1,2,\ldots,n$, $A=[a_1, a_2, \ldots, a_n]\in\mathcal{R}^{m\times n}$, and
let $k=\mathrm{rank}(A)$.  Assume that the mean of each column $a_j$ is zero
and that each column has unit norm $\|a_j\| = 1$. Here and below, $\|\cdot\|$
denotes the Euclidean vector norm. Then the Pearson sample correlation matrix
$\mathrm{cor}(A)=A^TA$.  Note that under these assumptions the sample
correlation and covariance matrices are the same.  Section \ref{irlba} below
illustrates how to relax the unit norm and zero mean assumptions in practice.

@@

\begin{lemma}
Let $a_i$ and $a_j$, $1\le i,j\le n$ be any two columns of the matrix
$A$. Then
\begin{equation}\label{cordist}
\mathrm{cor}(a_i,a_j) = a_i^Ta_j = 1 - \|a_i - a_j\|^2/2.
\end{equation}
In particular, for a given correlation threshold $t$, $0<t<1$, if
$\|a_i - a_j\|^2 > 2(1-t)$ then
$\mathrm{cor}(a_i, a_j) < t$.
\end{lemma}
%%\begin{proof}
%%\begin{eqnarray*}
%%\|a_i - a_j\|^2 &=& (a_i - a_j)^T(a_i - a_j) \\
%%&=& a_i^T a_i + a_j^T a_j - 2a_i^T a_j \\
%%&=& \|a_i\|^2 + \|a_j\|^2 - 2a_i^T a_j \\
%%&=& 2 - 2a_i^T a_j.
%%\end{eqnarray*}
%%The lemma follows by solving for $a_i^Ta_j$.
%%\end{proof}

Equation \ref{cordist} equates the problem of finding pairs of highly
correlated column vectors from the matrix $A$ with a problem of finding paris
of vectors sufficiently close together. For example, correlation values larger
than 0.99 may only be associated with column vectors from the matrix $A$ whose
Euclidean distance is less than $\sqrt{0.02}$.  Vectors farther apart can not
meet the correlation threshold.

@@

\section{Efficient pruning using the truncated SVD}

Let $t$ be a given correlation threshold $0<t<1$, and let
$P\in\mathcal{R}^{n\times n}$ be a permutation matrix that orders then entries
of $v_1$ in increasing order.  It follows from Equations \ref{cordist} and
\ref{vdist} that for any $i,j=1,2,\ldots,n$, if $\sigma_1^2 |e_i^Tv_{1} -
e_j^Tv_{1j}|^2 > 2(1-t)$ then $\mathrm{cor}(a_i,a_j) < t$.  Let $\ell$ be the
longest run of successive points in $P v_1$ within the interval
$\sqrt{2(1-t)}/\sigma_1$. For example, Figure \ref{fig1} displays the entries
of $v_1$ for the small ``BicatYeast'' example (where $A\in\mathcal{R}^{70\times
419}$) from Section \ref{examples} in increasing order.  The lines in the plot
illustrate the interval associated with the correlation threshold $t=0.95$. The
quantity $\ell$ is obtained by rolling the interval across all the points in
the plot and counting the maximum number of points that lie within the
interval.  With respect to the ordering $P$, $\ell$ represents the biggest
index difference that paris of indices corresponding to correlated vectors
above the threshold can exhibit.

@@

Pairs of vectors below the correlation threshold that lie more than one index
difference apart relative to the permutation $P$ can be similarly pruned by
replacing $D^1$ with $D^j$, $j=2,3,\ldots,\ell$ in the above expressions.
Following this process, we can produce a well-pruned candidate set of possible
vector pairs that meet a given correlation threshold using $\ell$ matrix vector
products with matrices of order $n \times p$, where $p$ is chosen in practice
such that $p<<k$. Setting $p=10$ for the example shown in Figure
\ref{fig1} prunes all but $431$ possible vector pairs out of a total
$419(419-1)/2 = 87,571$ with $\ell=138$.

>>>>

The number of correlation coefficients grows quadratically with the number of
vectors. Simply computing all pairs of correlation coefficients and then
filtering out coefficients below a given threshold may not be computationally
feasable for high-dimensional data arising in modern genomics and other
applications. Considerable attention has been devoted to methods called
\emph{pruning methods} that cheaply prune away all but the most
highly-correlated pairs of vectors.  A variety of pruning methods have been
described for the thresholded correlation problem.  For example the TAPER and
related algorithms of Xiong et al. base pruning rules on the frequency of
mutually co-ocurring observations (rows) among vector pairs; see \cite{prune1,
prune2} and the references therein. Related methods only approximate a set
containing the most highly-correlated pairs using hashing
algorithms\cite{prune3}. A number of other methods arrive at approximate sets
of the most highly-correlated paris using dimensionality-reduction techniques
like the discrete Fourier transform\cite{timeseries}. The method of Wu, et
al.\cite{svd-similarity}, conceptually very similar to our approach, uses the
singular value decomposition (SVD) to find all pairs of close vectors with
respect to a distance metric (instead of correlation).  Our method finds all of
the most highly-correlated vector pairs with respect to a given threshold by
pruning along an intuitive path of decreasing variance defined by the data and
computed by the SVD.

Consider a real-valued data matrix $A$ consisting of $m$ observations of $n$
column vectors.  Denote the columns of $A$ as $a_j\in\mathcal{R}^m$ for
$j=1,2,\ldots,n$, $A=[a_1, a_2, \ldots, a_n]\in\mathcal{R}^{m\times n}$, and
let $k=\mathrm{rank}(A)$.  Assume that the mean of each column $a_j$ is zero
and that each column has unit norm $\|a_j\| = 1$. Here and below, $\|\cdot\|$
denotes the Euclidean vector norm. Then the Pearson sample correlation matrix
$\mathrm{cor}(A)=A^TA$.  Note that under these assumptions the sample
correlation and covariance matrices are the same.  Section \ref{irlba}
illustrates how to relax the unit norm and zero mean assumptions in practice.

@@

\begin{lemma}
\label{distlemma}
Let $a_i$ and $a_j$, $1\le i,j\le n$ be any two columns of the matrix
$A$. Then
\begin{equation}\label{cordist}
\mathrm{cor}(a_i,a_j) = a_i^Ta_j = 1 - \|a_i - a_j\|^2/2.
\end{equation}
In particular, for a given correlation threshold $t$, $0<t<1$, if
$\|a_i - a_j\|^2 > 2(1-t)$ then
$\mathrm{cor}(a_i, a_j) < t$.
\end{lemma}
%%\begin{proof}
%%\begin{eqnarray*}
%%\|a_i - a_j\|^2 &=& (a_i - a_j)^T(a_i - a_j) \\
%%&=& a_i^T a_i + a_j^T a_j - 2a_i^T a_j \\
%%&=& \|a_i\|^2 + \|a_j\|^2 - 2a_i^T a_j \\
%%&=& 2 - 2a_i^T a_j.
%%\end{eqnarray*}
%%The lemma follows by solving for $a_i^Ta_j$.
%%\end{proof}

Lemma \ref{distlemma} equates the problem of finding pairs of highly
correlated column vectors from the matrix $A$ with a problem of finding paris
of vectors sufficiently close together. For example, correlation values larger
than 0.99 may only be associated with column vectors from the matrix $A$ whose
Euclidean distance is less than $\sqrt{0.02}$.  Vectors farther apart can not
meet the correlation threshold.

@@

\section{Efficient pruning using truncated SVD}

Let $t$ be a given correlation threshold $0<t<1$, and let
$P\in\mathcal{R}^{n\times n}$ be a permutation matrix that orders then entries
of $v_1$ in increasing order.  It follows from Equations \ref{cordist} and
\ref{vdist} that for any $i,j=1,2,\ldots,n$, if $\sigma_1^2 |e_i^Tv_1 -
e_j^Tv_1|^2 > 2(1-t)$ then $\mathrm{cor}(a_i,a_j) < t$.  Let $\ell$ be the
longest run of successive points in $P v_1$ within the interval
$\sqrt{2(1-t)}/\sigma_1$. For example, Figure \ref{fig1} displays the entries
of $v_1$ for the small ``BicatYeast'' example (where $A\in\mathcal{R}^{70\times
419}$) from Section \ref{examples} in increasing order.  The lines in the plot
illustrate the interval associated with the correlation threshold $t=0.95$. The
quantity $\ell$ is obtained by rolling the interval across all the points in
the plot and counting the maximum number of points that lie within the
interval.  With respect to the ordering $P$, $\ell$ represents the biggest
index difference that paris of indices corresponding to correlated vectors
above the threshold can exhibit.

@@

Pairs of vectors below the correlation threshold that lie more than one index
difference apart relative to the permutation $P$ can be similarly pruned by
replacing $D^1$ with $D^j$, $j=2,3,\ldots,\ell$ in the above expressions.
Following this process, we can produce a well-pruned candidate set of vector
pairs that contains all pairs that meet a given correlation threshold using
$\ell$ matrix vector products with matrices of order $n \times p$, where $p$ is
chosen in practice such that $p<<k$. Setting $p=10$ for the example shown in
Figure \ref{fig1} prunes all but $431$ possible vector pairs out of a total
$419(419-1)/2 = 87,571$ with $\ell=138$.
